# Data

You can find a overview of the data that is included in the textbook package *fpp3* by running the following code:
```{r, eval = FALSE, echo = TRUE}
data(package = "fpp3")
```
```{r, eval = T, echo = FALSE, message=FALSE,warning=FALSE}
library(dplyr)
library(fpp3)
data(package = "fpp3")$results %>% as_tibble() %>% 
  select(Item, Title)  
```

These are avaiable when the fpp3 package is loaded, i.e.

```{r}
library(fpp3)
bank_calls
```

To load a specific data set explicitly in your R environment:

```{r}
data("bank_calls")
```

Other examples used in the videos and content on this website is available for download at [github.com/holleland/BAN430/tree/master/data](https://github.com/holleland/BAN430/tree/master/data). You should also be able to load them directly into R using the raw link:

```{r}
# CPI Norway
read.csv(
  "https://raw.githubusercontent.com/holleland/BAN430/master/data/CPI_norway.csv", sep = ";") %>% head()
```

For excel files it is a bit less convenient, because you will need to download the file. But you can let R do that for you (if you insist).

```{r}
loadExcel_url <- function(url) {
    temp_file <- tempfile(fileext = ".xlsx")
    download.file(url = url, destfile = temp_file, mode = "wb", quiet = TRUE)
    readxl::read_excel(temp_file)
}
loadExcel_url("https://github.com/holleland/BAN430/blob/master/data/NorwayEmployment_15-74years_bySex.xlsx?raw=true")
```

The code above will save the file temporary in your computers temporary folder and load it into R from there. You could also adjust the code so that it stores the file in your working directory by adjusting the function.

```{r, eval = FALSE}
temp_file <- paste0(getwd(), "/NorwayEmployment.xlsx")
```

But the easiest will maybe be to just download the files manually from github and save them in a data folder of your own.
