# R and Rstudio

In this course, we will be using R and Rstudio to e.g. visualize time series, estimate model parameters, forecast, etc. It is therefore essential to have some basic knowledge of how to write an R script and how to read in data and do some simple data manipulation for preparing the data for different time series analysis. Hopefully, most of you have some experience with R and Rstudio before. If you have, this will be a short recap, if not this will be a very short introduction covering the most basic operations.

### Installing R and Rstudio {.unnumbered}

1.  Install R:
    -   Go to: [cran.uib.no](https://cran.uib.no/)
    -   Press download R for Linux/MacOS/Windows
    -   Press base
    -   Download R-4.x.x for Linux/MacOS/Windows
    -   Run the installation using default options
2.  Install Rstudio
    -   Go to: [rstudio.com](https://www.rstudio.com/products/rstudio/)
    -   Select Rstudio desktop
    -   Press Download Rstudio desktop
    -   Select the Rstudio desktop with open source licence, which is free\
    -   Select the version for your operating system
    -   Run the installation using default settings
3.  Open Rstudio and check that it works (it should start without any error messages).
4.  Install the R-package of the book "fpp3".
    -   In Rstudio, select Tools -\> Install packages -\> write "fpp3" and make sure install dependencies is marked. Press Install. You can also run the following code in the console

```{r, eval = FALSE}
#| echo: true
install.packages("fpp3", dependencies = TRUE)
```

<iframe src="https://player.vimeo.com/video/745712614?h=ead0e1cf78" width="640" height="360" frameborder="0" allow="autoplay; fullscreen; picture-in-picture" allowfullscreen>

</iframe>

Other useful packages to install are

```{r installreadxl, eval = FALSE}
install.packages("tidyverse", dependencies = TRUE)
install.packages("readxl")
```

(will add other packages as we go)
