## Stationarity

```{r}
#| message: false
#| warning: false
#| echo: false
#| eval: true
library(lubridate)
library(tidyverse)
library(fpp3)
theme_set(
  theme_bw() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          strip.background = element_rect(fill = "white", color = "transparent"))
)

```

We return to Labarr's time series in 5 minutes series on youtube explaining stationarity. This is a key concept in time series. The video by Labarr gives a nice intuition for the concept. It separates between strong and weak stationarity. In this course, the weak stationarity condition will be the one we focus on, but it is nice to know that there are other definitions of stationarity as well. For a time series to be (weakly) stationary the mean and variance of the time series variable $Y_t$ should not depend on $t$. More formally, a time series is weakly stationary if

1) The expectation is constant: $\mu_t = \mu$
2) The variance is constant and finite: $\sigma_t^2 = \sigma^2 < \infty$
3) The covariance between two lagged variables only depend on the lag: $\gamma(r,s) = \gamma(h)$, where $h=|r-s|$.

<iframe width="560" height="315" src="https://www.youtube.com/embed/aIdTGKjQWjA" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

In the video, Labarr shows some examples of non-stationary time series which can be transformed to stationary time series by differencing. We will come back to this later in the course, when we talk about transformations and again when we study ARIMA models. We will also study examples of stationary time / non-stationary time series in the voluntary homework. You can also read about the stationarity in the [textbook](https://otexts.com/fpp3/stationarity.html), but for now, you may stop after the paragraph on stationarity. The book is less specific about their definition, but we will stick the weakly stationary definition above.


<!-- ## Autocorrelation -->
<!-- We will cover this in c -->



## White noise
A white noise series is a time series of uncorrelated observations with mean zero and finite variance. We will often write it as $Z_t$ where $Z_t \sim \mathrm{WN}(0,\sigma^2)$. The standard is that the series is uncorrelated, but we may require it to be independent (stronger assumption) and very often normally distributed. In that case we call it iid Gaussian white noise (iid = independent and identically distributed). Notation for this may be $Z_t \sim \text{iid}\,  \mathrm{WN}(0,\sigma^2)$. Let us generate a white noise series in R.

```{r}
library(fpp3)
set.seed(123) # To produce the same output
wn <- tsibble(
  t = 1:100,
  Z = rnorm(100, sd = 3), # Draws from N(0, 3^2) distribution
  index = "t"
)
wn %>%
  autoplot() +
  labs(title = "Gaussian White noise",
       subtitle = "iid WN(0,9)")
```

We can then plot the acf of the series:
```{r}
wn %>%
  ACF(Z) %>%
  autoplot() +
  labs(title="White noise ACF")
```
As you can see, all the correlations fall within the confidence bands. The series is uncorrelated. You can also find a similar example in the [textbook](https://otexts.com/fpp3/wn.html).

<!-- ## Exercises -->

<!-- 1.  -->
