# Exercises

```{r}
#| message: false
#| warning: false
#| echo: false
#| eval: true
library(lubridate)
library(tidyverse)
library(fpp3)
theme_set(
  theme_bw() + 
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          strip.background = element_rect(fill = "white", color = "transparent"))
)

```


1. Use the simulated data from [here](https://holleland.github.io/BAN430/3_adjustments.html#time-series-components-and-seasonal-adjustment). Compare the different methods with the truth from the simulation. Are there large differences?

<details>
<summary>Solution</summary>

</details>


2. Use the global\_economy data, select a country (e.g. Austria). Plot GDP, GDP per capita and GDP per capita inflation adjusted, GDP inflation adjusted.

<details>

<summary>Solution</summary>
```{r}
#| message: false
#| warning: false
library(fpp3)
library(tidyverse)
dat <- global_economy %>% 
  filter(Country == "Austria")
dat %>% autoplot(GDP)
dat %>% autoplot(GDP/Population)
dat %>% autoplot(GDP/Population * 100/CPI)
dat %>% autoplot(GDP * 100/CPI)

```
</details>

3. In the global\_economy data set, the CPI has a reference year of 2010. Do the necessary changes to inflation adjust GDP per capita with 1990 as reference year. 

<details>

<summary>Solution</summary>

```{r}
library(fpp3)
library(tidyverse)
dat <- global_economy %>% 
  filter(Country == "Austria")

# -- Extracting the CPI in 1990: --
cpi1990 <- dat %>% filter(Year ==1990) %>% pull(CPI)

# -- Transforming such that CPI1990 is 100 in 1990: --
dat <- dat %>% mutate(CPI1990 = CPI / cpi1990 * 100)

# -- Plotting Inflation adjusted GDP per capita: --
dat %>% autoplot(GDP/Population * 100/CPI1990) + 
  labs(y = "Inflation adjusted GDP per capita (1990 US$)") +
  geom_line(aes(y= GDP/Population * 100/CPI), col = 2)

# -- Comparing the two CPIs: --
dat %>% 
  pivot_longer(cols = c(CPI,CPI1990)) %>%
  ggplot(aes(x = Year, y = value, col = name)) + geom_line()+
  geom_hline(yintercept = 100) +
  scale_color_manual(values = c("red","blue"))+
  geom_segment(x = 1990,xend = 1990, y = -Inf, yend =100, lty = 2, col = "blue")+
  geom_segment(x = 2010,xend = 2010, y = -Inf, yend =100, lty = 2, col = "red")+
  labs(title = "Differences between CPI with reference year 1990 and 2010",
       y = "Consumer Price Index") +
  theme(legend.title = element_blank())

```
</details>

4. Use the quantmod package to download data for another stock. Will a log transformation do a good job for this as well? 

5. Use the guerrero feature to select a $\lambda$ for the Box-Cox transformation on the data from the previous exercise. 









