## Forecasting volatility in R

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
theme_set(theme_bw() + theme(panel.grid.minor = element_blank(), 
                             panel.grid.major = element_blank(),
                             strip.placement = "outside",
                             strip.background = element_rect(fill = "white", 
                                                             color = "transparent")))
library(rugarch)
library(tidyquant) # Package getting stock data
library(tidyverse)
library(fpp3)
MSFT <- tq_get("MSFT", 
               get = "stock.prices",
               from = as.Date("2015-01-01")) %>% 
  select(date, close) %>% 
  mutate(tradingday = row_number()) %>% 
  as_tsibble(index = tradingday) %>% 
  mutate(return = difference(close)/close)
```

The fable package that we are using for everything else does not cover volatility modeling. We could maybe trick it, by using the ARMA representation of a GARCH model and back-transforming the estimates to the GARCH parameters, but this is not very convenient. It is also useful for you as students to see something other than the tidy packages. 

We use a package called [rugarch](https://cran.r-project.org/web/packages/rugarch/rugarch.pdf) for fitting and forecasting GARCH models. You find the vignette *Introduction to the rugarch package* [here](https://cran.r-project.org/web/packages/rugarch/vignettes/Introduction_to_the_rugarch_package.pdf). Install and load the package.
```{r, eval = FALSE, echo = T}
install.packages("rugarch")
library(rugarch)
``` 

We continue with the Microsoft closing price returns as example. We start by specifying the model. 
```{r, echo = T}
model <- ugarchspec(variance.model = list(model = "sGARCH", 
                                          garchOrder = c(3,1)),
                    mean.model = list(armaOrder = c(0,0), include.mean = FALSE),
                    distribution.model = "norm")
fit <- ugarchfit(model, 
                 data = MSFT %>% filter(!is.na(return)) %>% pull(return)
)
sigma(fit)
plot(ugarchforecast(fit, n.ahead = 200))
```

## References {.unnumbered}

- Engle, R. F. (1982). Autoregressive conditional heteroscedasticity with estimates of the variance of United Kingdom inflation. Econometrica: Journal of the econometric society, 987-1007.
- McNeil, A. J., Frey, R., & Embrechts, P. (2005). Quantitative risk management: concepts, techniques and tools-revised edition. Princeton university press.
