---
format:
  pdf:
    papersize: a4
    geometry:
      - margin=3cm
    toc: false
---

## Exponential smoothing {.unnumbered}

```{r}
#| message: false
#| warning: false
#| echo: false
#| eval: true
library(lubridate)
library(tidyverse)
library(fpp3)
theme_set(
  theme_bw() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          strip.background = element_rect(fill = "white", color = "transparent"))
)
```

📄 [Print-friendly PDF version](https://raw.githubusercontent.com/holleland/BAN430/master/pdfs/Lab_ets.pdf)

Four series, each from a different part of the Norwegian economy: how many people live here, what a US dollar costs in kroner, what it costs a bank to borrow overnight from Norges Bank, and how many nights people spent in Norwegian hotels. For each one, split into training and test data, fit a few candidate `ETS()` specifications, forecast, and check against the test set what actually helped.

### The data

```{r}
#| eval: false
install.packages("remotes")
remotes::install_github("holleland/ban430data")
```

```{r}
library(fpp3)
library(ban430data)
data(decomposition)
data(ets_lab)
```

| Object | Series | Contents |
|:---|:---|:---|
| `population_no` | 1 | Norway's national population at the start of each quarter, 1997 Q4 → today |
| `exchange_usdnok` | 2 | average monthly USD/NOK spot exchange rate, 2010 → today |
| `policy_rate_no` | 3 | Norges Bank's key policy rate, monthly, 2010 → today |
| `hotel_guestnights` | 4 | monthly guest nights at hotels and similar accommodation, all of Norway, 1986 → today |

#### Cheat sheet: `ETS()` syntax

| Term | Syntax |
|:---|:---|
| Simple exponential smoothing | `ETS(y ~ error("A") + trend("N") + season("N"))` |
| Holt's linear trend | `ETS(y ~ error("A") + trend("A") + season("N"))` |
| Damped trend | `ETS(y ~ error("A") + trend("Ad") + season("N"))` |
| Holt-Winters, additive season | `ETS(y ~ error("A") + trend("A") + season("A"))` |
| Holt-Winters, multiplicative season | `ETS(y ~ error("M") + trend("A") + season("M"))` |
| Let `ETS()` choose everything | `ETS(y)` |
| Fit | `model(name = ETS(...))` |
| Inspect a fitted model | `report(fit)` or `fit %>% select(name) %>% report()` |
| Forecast a known window | `forecast(fit, new_data = test)` |
| Accuracy vs. the actuals | `accuracy(fc, full_data)` |

------------------------------------------------------------------------

### Series 1: Norwegian population

Quarterly, 1997 Q4 → today.

```{r}
train_pop <- population_no %>% filter(quarter < yearquarter("2023 Q3"))
test_pop  <- population_no %>% filter(quarter >= yearquarter("2023 Q3"))
```

Fit a few candidate `ETS()` specifications on `train_pop`, forecast `test_pop`, and compare their accuracy.

<details>

<summary>💡 Code hint</summary>

``` r
fit_pop <- train_pop %>%
  model(
    SES    = ETS(population ~ error("A") + trend("N")  + season("N")),
    Holt   = ETS(population ~ error("A") + trend("A")  + season("N")),
    Damped = ETS(population ~ error("A") + trend("Ad") + season("N")),
    Auto   = ETS(population)
  )
fit_pop %>% forecast(new_data = test_pop) %>% accuracy(population_no) %>%
  select(.model, RMSE, MASE) %>% arrange(RMSE)
```

</details>

Print `report()` for the automatically selected model (`fit_pop %>% select(Auto) %>% report()`) and read off which components it actually contains.

**Q:** Which components — trend, damping, season — actually earned their place on the test set, and which just added complexity without improving the forecast?

------------------------------------------------------------------------

### Series 2: the USD/NOK exchange rate

Monthly, 2010 → today.

```{r}
train_fx <- exchange_usdnok %>% filter(month < yearmonth("2025-09"))
test_fx  <- exchange_usdnok %>% filter(month >= yearmonth("2025-09"))
```

Fit a few candidate `ETS()` specifications on `train_fx`, forecast `test_fx`, and compare their accuracy.

<details>

<summary>💡 Code hint</summary>

``` r
fit_fx <- train_fx %>%
  model(
    SES    = ETS(rate ~ error("A") + trend("N")  + season("N")),
    Holt   = ETS(rate ~ error("A") + trend("A")  + season("N")),
    Auto   = ETS(rate),
    Naive  = NAIVE(rate)
  )
fit_fx %>% forecast(new_data = test_fx) %>% accuracy(exchange_usdnok) %>%
  select(.model, RMSE, MASE) %>% arrange(RMSE)
```

</details>

Print `report()` for the automatically selected model and for `Naive`'s closest competitor — what do their fitted parameters have in common?

**Q:** Did adding a trend component help here, hurt, or make no real difference? What does that tell you about which component actually matters for this series?

------------------------------------------------------------------------

### Series 3: Norges Bank's key policy rate

Monthly. Train on data only through the end of 2023, and forecast the two years after.

```{r}
train_rate <- policy_rate_no %>% filter(month < yearmonth("2024-01"))
test_rate  <- policy_rate_no %>% filter(month >= yearmonth("2024-01"), month < yearmonth("2026-01"))
```

Fit a couple of candidate `ETS()` specifications on `train_rate` and forecast `test_rate`.

<details>

<summary>💡 Code hint</summary>

``` r
fit_rate <- train_rate %>%
  model(
    Holt   = ETS(rate ~ error("A") + trend("A")  + season("N")),
    Damped = ETS(rate ~ error("A") + trend("Ad") + season("N"))
  )
fc_rate <- fit_rate %>% forecast(new_data = test_rate)
fc_rate %>% accuracy(policy_rate_no) %>% select(.model, RMSE, MASE)
```

</details>

Print `report()` for both models, then plot both forecasts against what actually happened:

``` r
fc_rate %>% autoplot(policy_rate_no %>% filter(month >= yearmonth("2022-01")), level = NULL)
```

**Q:** By how many percentage points do the two forecasts disagree two years out? What did the rate actually do? Which single component — trend or damping — is responsible for almost all of that gap?

------------------------------------------------------------------------

### Series 4: hotel guest nights

Monthly, 1986 → today. This is the one from the decomposition lecture.

```{r}
hotel <- hotel_guestnights %>% rename(guests = `Guest nights`)

train_hotel <- hotel %>% filter(yearmonth < yearmonth("2024-07"))
test_hotel  <- hotel %>% filter(yearmonth >= yearmonth("2024-07"))
```

Fit a few candidate `ETS()` specifications on `train_hotel`, forecast `test_hotel`, and compare their accuracy.

<details>

<summary>💡 Code hint</summary>

``` r
fit_hotel <- train_hotel %>%
  model(
    SES     = ETS(guests ~ error("A") + trend("N")  + season("N")),
    HW_add  = ETS(guests ~ error("A") + trend("A")  + season("A")),
    HW_mult = ETS(guests ~ error("M") + trend("A")  + season("M")),
    Auto    = ETS(guests)
  )
fit_hotel %>% forecast(new_data = test_hotel) %>% accuracy(hotel) %>%
  select(.model, RMSE, MASE) %>% arrange(RMSE)
```

</details>

Print `report()` for the automatically selected model.

**Q:** Adding a season component clearly matters here — but does it matter *how* you add it? Look back at the full series plot: is there anything around early 2020 that might explain why an additive vs. multiplicative season makes such a difference?

------------------------------------------------------------------------

### Wrap-up discussion

- For each of the four series: which ETS components (trend, damping, season, and the choice between additive/multiplicative) actually mattered, and which ones didn't move the needle? Is there a series where none of the extra components beat simple exponential smoothing?
- In how many of your four series did `ETS()`'s own automatic pick actually win on the test set? Would you trust automatic selection for a model you had to actually deploy — or would you always want to check it against a held-out test set, and read its `report()`, first?
- The policy-rate series trains right up to the peak of a rapid change. Real deployed forecasting systems don't get to choose such a convenient cutoff — how would you *know*, in practice, that you were sitting at a turning point where an undamped trend was about to go badly wrong?
- Which of today's four series would you be comfortable handing a fully automatic `ETS()` pipeline for, with no human checking the forecast before it's used? Which would you not — and what would you want a human to check first?
