## Time series decomposition {.unnumbered}

```{r}
#| message: false
#| warning: false
#| echo: false
#| eval: true
library(lubridate)
library(tidyverse)
library(fpp3)
theme_set(
  theme_bw() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          strip.background = element_rect(fill = "white", color = "transparent"))
)
```

📄 [Print-friendly PDF version](Lab_decomposition.pdf)

### Before you start

This is an in-class lab: work through the material below at your own pace (in pairs or small groups is fine) — I'll walk around to help, and we'll stop along the way to discuss as a group.

This builds on the moving averages, classical decomposition, X11/SEATS and STL material from the [decomposition chapter](3_timeseriesdecomposition.qmd). Today you apply it: decompose real series into trend-cycle, seasonal and remainder components, compare methods, and interpret what you get.

```{r}
library(fpp3)
library(tidyverse)
```

#### Cheat sheet: decomposition methods

| Method | Syntax |
|:-----------------------------------|:-----------------------------------|
| Classical | `model(classical_decomposition(variable, type = "additive"))` |
| X11 | `model(X_13ARIMA_SEATS(variable ~ x11()))` |
| SEATS | `model(X_13ARIMA_SEATS(variable ~ seats()))` |
| STL | `model(STL(variable ~ trend(window = ...) + season(window = ...), robust = ...))` |

After fitting, use `components()` to extract the trend-cycle/seasonal/remainder columns, then `autoplot()` to plot them.

------------------------------------------------------------------------

### Task 1: Offshore wind power

Use the `windpower` data from `ban430data` (the same one from the [Time series graphics lab](Lab_timeseriesgraphics.qmd)) again — this time for decomposition.

```{r}
#| eval: false
# install.packages("remotes")
remotes::install_github("holleland/ban430data")
library(ban430data)
data(windpower)
```

a)  Create a tsibble containing the **daily** wind power data from **Sørlig Nordsjø 2** only (aggregate `windpower` from hourly to daily with the mean, same as in the graphics lab).

    <details>

    <summary>💡 Hint</summary>

    ``` r
    wp_day <- windpower %>%
      filter(Place == "Sørlig Nordsjø 2") %>%
      index_by(date = as_date(datetime)) %>%
      summarise(powerprod = mean(powerprod, na.rm = TRUE))
    ```

    </details>

b)  Decompose the series into trend-cycle $T_t$, season $S_t$ and remainder $R_t$, using a suitable decomposition method. Which method(s) from the cheat sheet above are *not* usable here, and why? Why did you choose the method you did?

    <details>

    <summary>💡 Hint</summary>

    ``` r
    wp_day %>%
      model(STL(powerprod)) %>%
      components() %>%
      autoplot()
    ```

    </details>

c)  We can also use **STL features** to assess and compare *how* seasonal or trending a series is rather than just eyeballing a decomposition (see [Section 4.3 of the textbook](https://otexts.com/fpp3/stlfeatures.html)). Aggregate `windpower` to **monthly** data for **both** locations (like `wp_month` in the graphics lab), then run:

    ``` r
    windpower %>%
      group_by_key() %>%
      index_by(yearmonth = yearmonth(datetime)) %>%
      summarise(powerprod = mean(powerprod, na.rm = TRUE)) %>%
      features(powerprod, feat_stl)
    ```

    Which location has the stronger `seasonal_strength_year`? What about `trend_strength`?

------------------------------------------------------------------------

### Task 2: Norwegian wholesale and retail sales index

The data is a monthly index for *Retail trade, except of motor vehicles and motorcycles*, from [Statistics Norway (table 07129)](https://www.ssb.no/statbank/table/07129) (Jan 2000 onwards). It's the `wholesale_raw_lab` object from `ban430data`.

```{r}
library(ban430data)
data(decomposition)

wholesale <- wholesale_raw_lab
head(wholesale, 3)
```

a)  Convert the `month` column to a `yearmonth` type and turn `wholesale` into a tsibble.

    <details>

    <summary>💡 Hint</summary>

    ``` r
    wholesale <- wholesale_raw_lab %>%
      rename(index = `wholesale and retail sales index`) %>%
      mutate(month = yearmonth(month)) %>%
      as_tsibble(index = month)
    ```

    </details>

b)  Make a time plot.

    <details>

    <summary>💡 Hint</summary>

    ``` r
    wholesale %>% autoplot(index)
    ```

    </details>

c)  Decompose the series using the classical, X11, SEATS and STL methods. Can you detect any prominent differences between the methods?

    <details>

    <summary>💡 Hint</summary>

    ``` r
    wholesale %>%
      model(
        classical = classical_decomposition(index, type = "additive"),
        x11       = X_13ARIMA_SEATS(index ~ x11()),
        seats     = X_13ARIMA_SEATS(index ~ seats()),
        stl       = STL(index)
      ) %>%
      components() %>%
      autoplot()
    ```

    </details>

d)  Try adjusting the trend and season windows of the STL (default values are 21 and 11 respectively). What happens?

    <details>

    <summary>💡 Hint</summary>

    ``` r
    wholesale %>%
      model(STL(index ~ trend(window = ...) + season(window = ...))) %>%
      components() %>%
      autoplot()
    ```

    </details>

e)  Using X11 or SEATS - the "statistics agency methods" from the cheat sheet - plot your seasonally-adjusted time series.

    <details>

    <summary>💡 Hint</summary>

    ``` r
    wholesale %>%
      model(seats = X_13ARIMA_SEATS(index ~ seats())) %>%
      components() %>%
      autoplot(season_adjust)
    ```

    </details>

f)  Statistics Norway publishes its own official seasonally-adjusted version of this exact series - it's the `wholesale_sa` object in `ban430data` (column `index_sa`; same table, same industry filter, official `VerdiinSesong` content code). Plot it alongside your series from (e). Do they agree?

    <details>

    <summary>💡 Hint</summary>

    ``` r
    own_sa <- wholesale %>%
      model(seats = X_13ARIMA_SEATS(index ~ seats())) %>%
      components() %>%
      as_tibble() %>%
      select(month, own = season_adjust)

    official_sa <- wholesale_sa %>%
      rename(month = yearmonth, official = index_sa)

    own_sa %>%
      left_join(official_sa, by = "month") %>%
      pivot_longer(c(own, official), names_to = "source", values_to = "index") %>%
      as_tsibble(index = month, key = source) %>%
      autoplot(index)
    ```

    </details>

g)  Using your method of choice, plot the detrended series.

    <details>

    <summary>💡 Hint</summary>

    ``` r
    wholesale %>%
      model(STL(index)) %>%
      components() %>%
      mutate(detrended = index - trend) %>%
      autoplot(detrended)
    ```

    </details>

h)  *Optional*: Implement your own additive classical decomposition by hand (moving averages + seasonal averaging), and compare to `classical_decomposition()`. (See [Chapter 3 exercises, exercise 6](3_exercises.qmd) if you want to check your approach against a worked solution.)

------------------------------------------------------------------------

### Wrap-up discussion

For both series today: did the different decomposition methods roughly agree, or did they tell noticeably different stories? When might you prefer STL over classical decomposition, or vice versa? What made X11/SEATS unsuitable for the daily wind power data?